Solving the Nonstationary Richards Equation\\ With Adaptive $hp$-FEM

### 2011-04-27 paper.tex
Last changes.

The Richards equation is obtained by combining the Darcy-Buckingham's law
with the law of conservation of mass. Its numerical solution 
is the subject of an intense research since the early 1970s, beginning with Neuman's 
finite element solution of unsteady dam seepage~\cite{neuman1}, Celia's mass 
conservative modified Picard iteration~\cite{celia} method in the early 1990s, 
 the recent Kees's work with focus on stabilized finite element 
approximations~\cite{kees,kees2,kees3}.
Numerical methods for the Richards equation
can be tested on several simplified examples with known exact solutions 
such as \cite{tracy1,tracy2}.

@@

One of the goals of this paper is to show that to certain extent, these problems 
can be avoided by resolving areas with steep solution gradients with adaptive 
finite element methods, so that 
the underlying reason for the nonphysical oscillations ceases to exist. In particular,
we focus on the application of adaptive $hp$-FEM, a modern version of the finite element 
method that is capable of extremely fast, exponential convergence \cite{babuska1}
(subject to certain assumptions on the solution of the problem which do not exclude
singular solutions and solutions with steep gradients).

@@

In Appendix we present weak forms for the Richards equation and their solution 
via the Picard's and Newton's method, describe in more detail some particular aspects on adaptive 
$hp$-FEM that are relevant for computations performed in this paper. We also briefly 
introduce the open source adaptive $hp$-FEM library HERMES \cite{hermes} that was used to 
generate numerical results for this paper.

@@

A relation between the water content and the pressure head is called 
{\em retention curve}. Its most commonly used analytical formula is 
the van Genuchten's equation~\cite{vangenuchten}:

@@

Note that having an exact solution allows us to calculate the approximation 
error very accurately, and thus compare fairly the performance of different methods. 
The problem will be solved using adaptive $hp$-FEM, adaptive $h$-FEM with 
quadratic elements $(p = 2)$, and adaptive $h$-FEM with linear elements $(p = 1)$
using the open-source finite element library HERMES\footnote{http://hpfem.org/hermes}. 
The library will be described in more 
detail in Section \ref{sec:hermes}.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[height=4.5cm]{mesh_init_h1.png}
\end{center}
\vspace{-6mm}
\caption{Initial mesh for $h$-FEM with linear elements (24843 DOF).}
\label{fig:init-meshes-3}
\end{figure}
In this and all consequent examples we use the implicit second-order 
Crank-Nicolson method for time integration.
In Figs. \ref{fig:init-meshes-1} - \ref{fig:init-meshes-3} 
different colors are used to distinguish between 
different polynomial degrees of finite elements. All elements in these computations 
are quadrilaterals. Notice in Fig. \ref{fig:init-meshes-1} that adaptive $hp$-FEM employs
anisotropic (i.e., directionally different) polynomial degrees in finite 
elements -- this is when an element in the mesh contains two different colors. 
The steady-state solution is shown in Fig. \ref{fig:steady-state}.

@@

This example is motivated by Kees' recent publication~\cite{kees}.  It is
concerned with an unsteady seepage of groundwater into an initially dry lysimeter with 
a side bottom orifice. The domain of interest is a rectangle $(8.0, 0)  \times (0, 6.5)$ m$^2$ 
bounded by $\Gamma_{1...6}$ as shown in Fig. \ref{fig:kees-uloha}.

@@

The domain consists of two 50 cm thick surface layers and a low-permeable
clay block. The unsaturated material properties were obtained from~\cite{unsat-prop} 
and the saturated material properties (storativity) \cite{pech}. 
The first layer is a silt loam, and the second is a silty clay 
loam. Layer three is sandy clay loam and layer four silty clay again. 
For details on the material parameters see Fig.~\ref{fig:kees-uloha} 
and Table~\ref{Material-properties}. For the corresponding constitutive 
relations we use the van Genuchten's law~(\ref{vangenuchten-eq}) and Mualem's 
law~(\ref{mualem-int}).

@@

The reader can experiment with this example on his/her 
own, it is available in the HERMES code repository \cite{hermes-repo}
as example "richards-seepage-adapt".

@@

It is well known that the $h$-based form~(\ref{rich-final}) 
of the Richards equation is prone 
to errors in mass balance as well as to erroneous estimates of 
infiltration depth when inaccurate time integration is used (see, e.g., 
Celia et. al.~\cite{celia}). These errors are caused by improper discrete 
approximation of the time derivative term 
$C(h)\frac{\partial h}{\partial t}$ that originates in the mass (water content) derivative 
term $\frac{\partial \theta(h)}{\partial t}$. Kees et al.~\cite{kees-time-integ} have 
shown in the context of two-phase flow that higher-order time integration methods
can cure the mass balance problem.

@@

\section{Weak Formulation of the Richard's Equation
for the Newton's and Picard's Methods}\label{sec:9}

@@

Edge functions are higher-order polynomials
that are associated with mesh edges. On such 
an edge, the edge function has a nonzero trace,
and it is nonzero within the elements that shareHigher order time integration methods for two-phase flow
the edge (see Fig. \ref{fig:edge-fn}).

@@

The library can be employed to solve various PDE problems ranging 
from stationary linear equations to complex 
time-dependent nonlinear multiphysics PDE systems. The code repository \cite{hermes-repo}
accessible on the project home page contains many examples from various 
engineering and scientific areas including civil, mechanical, 
electrical and nuclear engineering, computational fluid dynamics, 
quantum chemistry, and hydrology.

@@

The next part of the User Documentation is the Tutorial that is split 
into six sections:

\begin{enumerate}
\item Linear Problems.
\item Automatic Adaptivity.
\item Nonlinear Problems.
\item Adaptivity for Time-Dependent Problems.
\item Miscellaneous Techniques.
\item Using Trilinos.
\end{enumerate}
The last two sections of the User Documentation contain 
a detailed description of about twenty benchmark problems
with known analytical solutions including convergence studies, 
and about twenty advanced examples.

@@

The user has to write a C++ application in order to utilize 
the library's functionality. For typical problems, this 
applications are extremely simple. Let us demonstrate this on 
the adaptive example that was used in Section \ref{sec:hpfem}.
The following code snippets are part of the main.cpp file 
from benchmark "smooth-iso" that can be retrieved from HERMES
code repository \cite{hermes-repo}.

One begins with specifying for each boundary marker whether the 
corresponding boundary condition is natural or essential. In 
this particular case, the entire boundary is Dirichlet and thus 
essential:

\begin{code}
// Boundary condition types.
BCType bc_types(int marker)
{
  return BC_ESSENTIAL;
}
\end{code}
In the next step, one defines boundary values for all 
essential boundary conditions. These may depend on the 
spatial position, possibly on time (via a global variable,
and for the example in Section \ref{sec:hpfem} 
they are zero:

\begin{code}
scalar essential_bc_values(int ess_bdy_marker, 
                           double x, double y)
{
  return 0;
}
\end{code}
In the next step the user enters the variational fomulation
of the PDE. The weak forms can be real or complex valued,
and there are two types of them -- {\em matrix forms} and
{\em vector forms}.

@@

\begin{itemize}
\item linear forms in linear problems,
\item residuals for nonlinear problems.
\end{itemize}
The example in Section \ref{sec:hpfem} 
solves the Poisson equation $-\Delta u = f$
with homogeneous Dirichlet boundary conditions,
and thus the corresponding matrix form is 
$$
\int_{\Omega} \nabla u \cdot \nabla v \, \mbox{d}x\mbox{d}y.
$$
This is translated into the code as follows:

\begin{code}
// Matrix form.
template<typename Real, typename Scalar>
Scalar bilinear_form(int n, double *wt, 
                     Func<Scalar> *u_ext[], 
                     Func<Real> *u, Func<Real> *v, 
                     Geom<Real> *e, 
                     ExtData<Scalar> *ext)
{
  return int_grad_u_grad_v<Real, Scalar>(n, wt, u, v);
}
\end{code}
Here, {\tt int\_grad\_u\_grad\_v} is a predefined macro 
that performs the numerical integration in this widely used
case. Without the macro, the code would read

\begin{code}
// Matrix form.
template<typename Real, typename Scalar>
Scalar bilinear_form(int n, double *wt, 
                     Func<Scalar> *u_ext[], 
                     Func<Real> *u, Func<Real> *v, 
                     Geom<Real> *e, 
                     ExtData<Scalar> *ext)
{
  Scalar val = 0;
  for (int i = 0; i < n, i++) {
    val += wt[i] * (u->dx[i] * v->dx[i] 
                    + u->dy[i] * v->dy[i]);
  }

  return val;
}
\end{code}
Here {\tt n} is the number of integration points in the mesh
element and {\tt wt} is an array of the corresponding integration
weights. Physical coordinates of the integration points can be 
accessed through the Geom structure along with other geometrical 
information such as normal or tangential vectors on edges, etc.
The variable {\tt u\_ext} contains the solution values and first and 
second partial derivatives at all integration points, and it is 
relevant for nonlinear problems only. The variables {\tt u} and 
{\tt v} contain the same information for the basis function {\tt u}
and test function {\tt v}. The struct {\tt ExtData} is used in 
more advanced application to provided additional data for the 
weak formulation.

The right-hand side of the weak formulation has the form 
$$
\int_{\Omega} fv\, \mbox{d}x\mbox{d}y.
$$
The function $f$ is defined via

\begin{code}
// Right-hand side function.
template<typename Real>
Real rhs(Real x, Real y)
{
  return 2*sin(x)*sin(y);
}
\end{code}
and the linear form using

\begin{code}
// Vector form.
template<typename Real, typename Scalar>
Scalar linear_form(int n, double *wt, 
                   Func<Scalar> *u_ext[], 
                   Func<Real> *v, Geom<Real> *e, 
                   ExtData<Scalar> *ext)
{
  return int_F_v<Real, Scalar>(n, wt, rhs, v, e);
}
\end{code}
Again the function {\tt int\_F\_v} is a shortcut and 
the full version of this function would be

\begin{code}
// Vector form.
template<typename Real, typename Scalar>
Scalar linear_form(int n, double *wt, 
                   Func<Scalar> *u_ext[], 
                   Func<Real> *v, Geom<Real> *e, 
                   ExtData<Scalar> *ext)
{
  Scalar val = 0;
  for (int i = 0l i < n; i++) {
    val += wt[i] * rhs(e->x[i], e->y[i]) * v->val[i];
  }

  return val;
}
\end{code}
Here the user can see how the physical coordinates of 
integration points are accessed via the {\tt Geom}
structure, and how one uses function values of the 
test function {\tt v} at integration points.

After entering the weak formulation, the problem 
can be solved. This is done using a standard sequence 
of steps. First, the mesh is loaded:

@@

\begin{code}
Mesh::refine_element(int id, int refinement = 0);
Mesh::convert_quads_to_triangles();
Mesh::convert_triangles_to_quads();
Mesh::refine_by_criterion(int (*criterion)(Element* e), 
                          int depth);
Mesh::refine_towards_vertex(int vertex_id, int depth);
Mesh::regularize(int n);
Mesh::unrefine_element(int id);
Mesh::unrefine_all_elements();
\end{code}
Next, one creates a finite element space:
\begin{code}
  // Create an H1 space with default shapeset.
  H1Space space(&mesh, bc_types, essential_bc_values, 
                P_INIT);
\end{code}
Here, {\tt P\_INIT} is an initial polynomial degree 
of mesh elements which currently can be an integer
between one and ten. Other admissible spaces in
addition to {\tt H1Space} are {\tt HcurlSpace},
{\tt HdivSpace} and {\tt L2Space}. In vector-valued
PDE problems, each solution component can have 
a different space.

After defining spaces for all solution components, 
one needs to register the weak forms in the {\tt WeakForm}
class:

\begin{code}
  // Initialize the weak formulation.
  WeakForm wf;
  wf.add_matrix_form(callback(bilinear_form), HERMES_SYM);
  wf.add_vector_form(callback(linear_form));
\end{code}
The number of weak forms that are added here 
can be arbitrary.

In typical non-adaptive computations one just needs 
to call the function {\tt solve\_linear()}:

\begin{code}
  // Solve the linear problem.
  Solution sln;
  solve_linear(&space, &wf, &sln, matrix_solver);
\end{code}
Through the variable {\tt matrix\_solver} the user tells 
his preference for the matrix solver. The options 
currently available include UMFPACK, PETSc, MUMPS, 
Trilinos and a variety of Python solvers available through 
an interface to Scipy \cite{scipy}. After the call to 
{\tt solve\_linear()} is finished, the variable {\tt sln}
contains a solution that can be either saved to a file
or visualized in HERMES directly using one of the {\tt ScalarView},
or {\tt VectorView} classes. Other visualization classes 
including {\tt MeshView}, {\tt BaseView}, etc. are available.

For adaptive computations, one needs to specify a refinement 
selector that manages the selection of element refinement 
candidates during adaptivity, as it was described in 
Subsection \ref{subsec:cand}. In the example in Section \ref{sec:hpfem}
we use an {\tt H1ProjBasedSelector}:

@@

\begin{code}
  // Initialize adaptivity parameters.
  AdaptivityParamType apt(ERR_STOP, NDOF_STOP, 
                          THRESHOLD, STRATEGY, 
                          MESH_REGULARITY);
\end{code}
The user-defined constant {\tt ERR\_STOP} defines the level of 
relative error estimate in percent when the computationa will stop. 
{\tt NDOF\_STOP} is the threshold for the number of degrees of freedom 
when the adaptivity will stop. The other two parameters, {\tt THRESHOLD}
and {\tt STRATEGY} are determining the behavior of the adaptivity 
algorithm and their meaning can be found in the User Documentation.
{\tt MESH\_REGULARITY} determines the maximum level of hanging nodes
in the mesh. Regular meshes are not supported in HERMES due to their 
notoriously bad performance. By specifying {\tt MESH\_REGULARITY = -1},
one allows arbitrary-level hanging nodes (cf. Subsection 
\ref{subsec:hang}).

For a typical linear or nonlinear problem the last step is to call the 
function call {\tt solve\_linear\_adapt()}:

\begin{code}
  // Adaptivity loop.
  Solution *sln = new Solution();
  Solution *ref_sln = new Solution();
  ExactSolution exact(&mesh, exact_sol);
  WinGeom* swg = new WinGeom(0, 0, 440, 350);
  WinGeom* mwg = new WinGeom(450, 0, 400, 350);
  bool verbose = true;    
  solve_linear_adapt(&space, &wf, HERMES_H1_NORM, sln, 
                     matrix_solver, ref_sln, 
                     &selector, &apt, swg, mwg, 
                     verbose, &exact_sol);
\end{code}
Here, {\tt sln} and {\tt ref\_sln} are the 
solution pair on the coarse and globally refined 
meshes, cf. Subsection \ref{subsec:cand}.
{\tt ExactSolution} is an optional parameter to be used
if the exact solution to the problem is know (as
it is known for the problem in Section \ref{sec:hpfem}).
The parameter {\tt H2D\_H1\_NORM} specifies that the
adaptivity algorithm will be projecting the fine mesh
solution on the coarse mesh using the $H^1$-norm. 
Other possible norms are {\tt H2D\_HCURL\_NORM},
{\tt H2D\_HDIV\_NORM} and {\tt H2D\_L2\_NORM}.
By setting the Boolean variable {\tt verbose} to either 
true or false, the user specifies whether he is 
interested in the output of adaptivity data, such as 
error estimates, numbers of degrees of freedom in the 
coarse and fine mesh solutions, etc. The variables
{\tt swg} and {\tt mwg} define geometries of 
visualization windows. By setting them to NULL, the
user disables visualization inside the {\tt solve\_linear\_adapt()}
function. The last optional parameter {\tt exact\_sol}
contains the exact solution (if available). If this 
parameter is provided, then HERMES will also measure
error wrt. exact solution in addition to its error
estimate.

For vector-valued problems, the 
user can replace single variables such as 
the space, projection norm, solution, reference solution, 
or views with Tuples.

Users who have more advanced applications can copy and 
edit the source code of the function {\tt solve\_linear\_adapt()}
according to their needs. For more details we refer to the 
Sphinx tutorial to HERMES that is available on the 
project home page \cite{hermes}. The tutorial also covers more advanced 
examples such as the solution of multiphysics coupled problems
using adaptive $hp$-FEM on dynamical meshes.

@@

\begin{figure*}[!ht]
\begin{center}
\includegraphics[height=8cm]{agros.png}
\end{center}
\vspace{-4mm}
\caption{Sample snapshot of AGROS on Mac OS X. An electromagnetic 
         simulation is shown, but a Richard's module is in development.}
\label{fig:agros}
\end{figure*}

@@

The first author gratefully acknowledges the financial support
of the Grant Agency of the Academy of Sciences of the Czech 
Republic under Grant No. IAA100760702. The second author
was supported with a Grant of the Czech Ministry of Agriculture No.
QH 91247. The third author's research 
was supported by the LC06040 research center project.

@@

\bibitem{kuraz-jcam}  M. Kuraz, P. Mayer, M. Leps, D. Trpkosova: An Adaptive Time Discretization of the Classical and the Dual Porosity Model of Richards' Equation, Journal of Computational and 
Applied Mathematics 233 (2010), 3167-3177

@@

\bibitem{miller} C. T. Miller, G. A. Williams, C. T. Kelley, M. D. Tocci: Robust Solution of Richards' Equation for Nonuniform Porous Media, Water Resour. Res. 34(10) (1998), 2599-2610.

@@

\bibitem{sdirk} F. Ismail, R.A. Al-Khasawneh, M. Suleiman: Embedded Singly Diagonally Implicit Runge-Kutta-Nystr\'om General Method (3,4) in (4,5) for Solving Second Order IVPs, International Journal of Applied Mathematics 37(2007)

>>>>

The Richards equation is obtained by combining the Darcy-Buckingham's law
with the law of conservation of mass. Its numerical solution 
is the subject of an intense research since the early 1970s, beginning with Neuman's 
finite element solution of unsteady dam seepage~\cite{neuman1}. 
%Celia's mass 
%conservative modified Picard iteration~\cite{celia} method in the early 1990s, 
% the recent Kees's work with focus on stabilized finite element 
%approximations~\cite{kees,kees2,kees3}.
Numerical methods for the Richards equation
can be tested on several simplified examples with known exact solutions 
such as \cite{tracy1,tracy2}.

@@

One of the goals of this paper is to show that to a certain extent, these problems 
can be avoided by resolving areas with steep solution gradients with adaptive 
finite element methods, so that 
the underlying reason for the nonphysical oscillations ceases to exist. In particular,
we focus on the application of adaptive $hp$-FEM, a modern version of the finite element 
method that is capable of extremely fast, exponential convergence \cite{babuska1}
(subject to certain assumptions on the solution of the problem which do not exclude
singular solutions and solutions with steep gradients).

@@

In Appendix A we present weak forms for the Richards equation and their solution 
via the Picard's and Newton's methods, describe in more detail some particular 
aspects on adaptive 
$hp$-FEM that are relevant for computations performed in this paper. In Appendix B we
briefly introduce the open source adaptive $hp$-FEM library HERMES \cite{hermes} 
that was used to generate numerical results for this paper.

@@

A relation between the water content and the pressure head is called 
a {\em retention curve}. Its most commonly used analytical formula is 
the van Genuchten's equation~\cite{vangenuchten}:

@@

Note that having an exact solution allows us to calculate the approximation 
error very accurately, and thus compare fairly the performance of different methods. 
The problem will be solved using adaptive $hp$-FEM, adaptive $h$-FEM with 
quadratic elements $(p = 2)$, and adaptive $h$-FEM with linear elements $(p = 1)$
using the open-source finite element library HERMES\footnote{http://hpfem.org/hermes}. 
The library will be described in more detail in Appendix B.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[height=4.5cm]{mesh_init_h1.png}
\end{center}
\vspace{-6mm}
\caption{Initial mesh for $h$-FEM with linear elements (24843 DOF).}
\label{fig:init-meshes-3}
\end{figure}
In this and all subsequent examples we use the implicit second-order 
Crank-Nicolson method for time integration.
In Figs. \ref{fig:init-meshes-1} - \ref{fig:init-meshes-3} 
different colors are used to distinguish between 
different polynomial degrees of finite elements. All elements in these computations 
are quadrilaterals. Notice in Fig. \ref{fig:init-meshes-1} that adaptive $hp$-FEM employs
anisotropic (i.e., directionally different) polynomial degrees in finite 
elements -- this is when an element in the mesh contains two different colors. 
The steady-state solution is shown in Fig. \ref{fig:steady-state}.

@@

This example was used previously to study discretizations for Richards' equation 
\cite{forsyth,kees}. It is
concerned with an unsteady seepage of groundwater into an initially dry lysimeter with 
a side bottom orifice. The domain of interest is a rectangle $(8.0, 0)  \times (0, 6.5)$ 
m$^2$ bounded by $\Gamma_{1...6}$ as shown in Fig. \ref{fig:kees-uloha}.

@@

The domain consists of two 50 cm thick surface layers and a low-permeability
clay block. The unsaturated material properties were obtained from~\cite{unsat-prop} 
and the saturated material properties (storativity) \cite{pech}. 
The first layer is a silt loam, and the second is a silty clay 
loam. Layer three is sandy clay loam and layer four silty clay again. 
For details on the material parameters see Fig.~\ref{fig:kees-uloha} 
and Table~\ref{Material-properties}. For the corresponding constitutive 
relations we use the van Genuchten's law~(\ref{vangenuchten-eq}) and Mualem's 
law~(\ref{mualem-int}).

@@

Readers can obtain this example as "richards-seepage-adapt" from the 
HERMES code repository in order to reproduce these results or conduct 
additional experiments \cite{hermes-repo}.

@@

It is well known that the $h$-based form~(\ref{rich-final}) 
of the Richards equation is prone 
to errors in mass balance as well as to erroneous estimates of 
infiltration depth when inaccurate time integration is used (see, e.g., 
Celia et. al.~\cite{celia}). These errors are caused by improper discrete 
approximation of the time derivative term 
$C(h)\frac{\partial h}{\partial t}$ that originates in the mass (water content) derivative 
term $\frac{\partial \theta(h)}{\partial t}$. Kees et al.~\cite{kees-time-integ} have 
shown how to solve the mass balance problem for higher-order time discretizations as well.

@@

\section{Weak Formulation of the Richards Equation
for the Newton's and Picard's Methods}\label{sec:9}

@@

Edge functions are higher-order polynomials
that are associated with mesh edges. On such 
an edge, the edge function has a nonzero trace,
and it is nonzero within the elements that share
the edge (see Fig. \ref{fig:edge-fn}).

@@

The library can be employed to solve various PDE problems ranging 
from stationary linear equations to complex 
time-dependent nonlinear multiphysics PDE systems. The code repository \cite{hermes-repo}
accessible on the project home page contains many examples from various 
engineering and scientific areas including civil, mechanical, 
electrical and nuclear engineering, computational fluid dynamics, 
quantum chemistry, and hydro\-lo\-gy.

@@

The next part of the User Documentation is the Tutorial that is split 
into 10 sections:

\begin{enumerate}
\item Linear Problems.
\item Nonlinear Problems.
\item Time-Dependent Problems.
\item Adaptive Solution of Linear Problems.
\item Adaptive Solution of Nonlinear Problems.
\item Adaptive Solution of Time-Dependent Problems.
\item Eigenvalue Problems.
\item FVM and DG.
\item Using Trilinos.
\item Miscellaneous Techniques.
\end{enumerate}
The last two sections of the User Documentation contain 
a detailed description of about twenty benchmark problems
with known analytical solutions including convergence studies, 
and about twenty advanced examples.

@@

In order to utilize the library's functionality, the user has 
to write a custom C++ application. For typical problems, these 
applications are very simple. Let us demonstrate this on 
the adaptive example that was used in Section \ref{sec:hpfem}.
The following code snippets are part of the main.cpp file 
from benchmark "lshape" that can be retrieved from HERMES
code repository \cite{hermes-repo}.

This problem comes with an exact solution (not shown here)
that is instantiated as follows,

\begin{code}
  CustomExactSolution exact_sln(&mesh);
\end{code}
The exact solution is used to define Dirichlet 
boundary conditions,

\begin{code}
  // Initialize boundary conditions
  DefaultEssentialBCNonConst 
    bc_essential(BDY_DIRICHLET, &exact_sln);
  EssentialBCs bcs(&bc_essential);
\end{code}
A predefined weak formulation is used for the Laplace equation,

\begin{code}
  // Initialize the weak formulation.
  DefaultWeakFormLaplace wf;
\end{code}
Weak forms in HERMES can be real or complex valued,
and there are two types of them -- {\em matrix forms} and
{\em vector forms}.

@@

\begin{itemize}
\item linear forms in linear problems,
\item residuals for nonlinear problems.
\end{itemize}
After that the problem 
is solved using a standard sequence 
of steps. First, the mesh is loaded
as follows,

@@

\begin{code}
Mesh::refine_element(int id, int refinement = 0);
Mesh::convert_quads_to_triangles();
Mesh::convert_triangles_to_quads();
Mesh::refine_towards_vertex(int vertex_id, int depth);
Mesh::regularize(int n);
Mesh::unrefine_element(int id);
Mesh::unrefine_all_elements();
\end{code}
Next, one creates a finite element space,
\begin{code}
  // Create an H1 space with default shapeset.
  H1Space space(&mesh, &bcs, P_INIT);
\end{code}
Here, {\tt P\_INIT} is an initial polynomial degree 
of mesh elements which currently can be an integer
between one and ten. Other admissible spaces in
addition to {\tt H1Space} are {\tt HcurlSpace},
{\tt HdivSpace} and {\tt L2Space}. In vector-valued
PDE problems, each solution component can have 
a different space.

After defining spaces for all solution components, 
one selects a matrix solver. The options 
currently available include MUMPS, PETSc, UMFpack, SUPERLU, 
Trilinos, and a variety of Python solvers available through 
an interface to Scipy \cite{scipy}.

For adaptive computations, one needs to specify a refinement 
selector that manages the selection of element refinement 
candidates during adaptivity, as it was described in 
\ref{subsec:cand}. In the example in Section 
\ref{sec:hpfem} we use an {\tt H1ProjBasedSelector}:

@@

\begin{code}
  // Initialize adaptivity parameters.
  AdaptivityParamType apt(ERR_STOP, NDOF_STOP, 
                          THRESHOLD, STRATEGY, 
                          MESH_REGULARITY);
\end{code}
The user-defined constant {\tt ERR\_STOP} defines the level of 
relative error estimate in percent when the computationa will stop. 
{\tt NDOF\_STOP} is the threshold for the number of degrees of freedom 
when the adaptivity will stop. The other two parameters, {\tt THRESHOLD}
and {\tt STRATEGY} are determining the behavior of the adaptivity 
algorithm and their meaning can be found in the User Documentation.
{\tt MESH\_REGULARITY} determines the maximum level of hanging nodes
in the mesh. Regular meshes are not supported in HERMES due to their 
notoriously bad performance. By specifying {\tt MESH\_REGULARITY = -1},
one allows arbitrary-level hanging nodes (cf. \ref{subsec:hang}).

The adaptivity loop follows. Its first step is to 
refine the mesh globally in both $h$ and $p$ and 
compute a reference solution,

\begin{code}
  // Construct globally refined reference 
  // mesh and setup reference space.
  Space* ref_space 
    = Space::construct_refined_space(&space);

  // Assemble the reference problem.
  info("Solving on reference mesh.");
  bool is_linear = true;
  DiscreteProblem* dp 
    = new DiscreteProblem(&wf, ref_space, is_linear);
  dp->assemble(matrix, rhs);

  // Solve the linear system of the reference problem.
  //  If successful, obtain the solution.
  Solution ref_sln;
  if(solver->solve()) Solution::vector_to_solution(
    solver->get_solution(), ref_space, &ref_sln);
  else error ("Matrix solver failed.\n");
\end{code}
In the next step the reference solution is projected on
the coarse mesh to obtain a low-order part
for error estimation,

\begin{code}
  // Project the fine mesh solution on the coarse mesh.
  Solution sln;
  info("Projecting reference solution on coarse mesh.");
  OGProjection::project_global(&space, 
    &ref_sln, &sln, matrix_solver);
\end{code}
With the solution pair we can proceed to error 
estimation,

\begin{code}
  // Calculate element errors and total error estimate.
  info("Calculating error estimate and exact error.");
  Adapt* adaptivity = new Adapt(&space);
  double err_est_rel 
    = adaptivity->calc_err_est(&sln, &ref_sln) * 100;
\end{code}
Finally, with an a-posteriori error estimate, we can 
adapt the mesh:

\begin{code}
  // If err_est too large, adapt the mesh.
  if (err_est_rel < ERR_STOP) done = true;
  else
  {
    info("Adapting coarse mesh.");
    done = adaptivity->adapt(&selector, THRESHOLD, 
                       STRATEGY, MESH_REGULARITY);

    // Increase the counter of performed 
    // adaptivity steps.
    if (done == false)  as++;
  }
  if (Space::get_num_dofs(&space) >= NDOF_STOP) 
    done = true;
\end{code}
For more details we refer to the 
Sphinx tutorial to HERMES that is available on the 
project home page \cite{hermes}.

@@

\begin{figure*}[!ht]
\begin{center}
\includegraphics[height=8cm]{agros.png}
\end{center}
\vspace{-4mm}
\caption{Sample snapshot of AGROS on Mac OS X. An electromagnetic 
         simulation is shown, but a Richards module is in development.}
\label{fig:agros}
\end{figure*}

@@

The first author gratefully acknowledges the financial support
of the Grant Agency of the Academy of Sciences of the Czech 
Republic under Grant No. IAA100760702. The second author
was supported with a Grant of the Czech Ministry of Agriculture No.
QH 91247.

@@

\bibitem{kuraz-jcam}  M. Kuraz, P. Mayer, M. Leps, D. Trpkosova: An Adaptive Time Discretization of the Classical and the Dual Porosity Model of Richards Equation, Journal of Computational and 
Applied Mathematics 233 (2010), 3167-3177

@@

\bibitem{miller} C. T. Miller, G. A. Williams, C. T. Kelley, M. D. Tocci: Robust Solution of Richards Equation for Nonuniform Porous Media, Water Resour. Res. 34(10) (1998), 2599-2610.

@@

%\bibitem{sdirk} F. Ismail, R.A. Al-Khasawneh, M. Suleiman: Embedded Singly Diagonally Implicit Runge-Kutta-Nystr\'om General Method (3,4) in (4,5) for Solving Second Order IVPs, International Journal of Applied Mathematics 37(2007)
